---
title: "Acute myeloid leukaemia (AML) data - code"
author: "Jackie Rao"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pheatmap)
library(RColorBrewer)
```

# Load data

Load data from cBioPortal via the package cBioPortalR and clean-up into a binary data matrix suitable for clustering with VICatMix.

```{r generateData}
library(cbioportalR)
library(maftools)
library(dplyr)

set_cbioportal_db(db = "public") #Set to public cBioPortal database

#Load AML mutations from TCGA
mutations <- get_mutations_by_study(study_id = "laml_tcga_pan_can_atlas_2018") 
#Format columns in line with MAF files
maf_conv_cols <- c("Hugo_Symbol", "Entrez_Gene_Id", "Sample_Key", "Patient_Key",
                   "Molecular_Profile_ID", "Tumor_Sample_Barcode", "Patient_Id",
                   "Study_Id", "Centre", "Mutation_Status", "Validation_Status",
                   "Start_Position", "End_Position", "Reference_Allele", "Protein_Change", 
                   "Variant_Classification", "NCBI_Build", "Variant_Type", "Keyword", "Chromosome",
                   "Tumor_Seq_Allele2", "RefSeq", "Protein_position", "Protein_end_position",
                   "t_alt_count", "t_ref_count", "normalAltCount", "normalRefCount") 
colnames(mutations) <- maf_conv_cols
mutations <- mutate(mutations, VAF = t_alt_count / (t_alt_count + t_ref_count))
mutations <- mutations[mutations$Mutation_Status == "Somatic",] #Keep only somatic mutations

#Convert to MAF 
laml <- read.maf(mutations, useAll = TRUE)
#Create mutation count matrix
mtx <- mutCountMatrix(laml, includeSyn = FALSE, countOnly = NULL, removeNonMutated = FALSE)

#Transpose mtx to have genes in columns and samples in row
mtx <- t(mtx)
#Convert counts to binary
mtx.b <- apply(mtx, 2, function(x) ifelse(x > 0, 1, x)) # So 0 = no, 1 =yes

mtx.b <- mtx.b[,-which(colSums(mtx.b) < 2)] #Remove mutations seen in less than 2 samples
mtx.b <- mtx.b[-which(rowSums(mtx.b) < 1),] #Remove patients with no mutations

```

# Run VICatMix

Perform the clustering, with 30 initialisations and 20 maximum clusters. Complete linkage and a 0.95 threshold for variable selection is used for the summarisation by default.

Uncomment the below lines to run the computation - these have been commented out so the Rmd file runs quickly.

```{r computation}

#set.seed(205) 
#library(VICatMix)
#library(foreach)
#library(doParallel)
#library(doRNG)
#registerDoParallel(30)

#amlclust_avg <- runVICatMixVarSelAvg(mtx.b, 20, 0.01, inits = 30, parallel = TRUE)

```

# Plots

## Heatmaps

Load data for plotting - this is the output from the above computation.
```{r load data}
#Comment out the following line if VICatMix has been run by the user
load("amlclust_avg.RData")
```

Create heatmaps with the 'pheatmap' package.
```{r heatmaps}
set.seed(205)
library(pheatmap)

rowshuffle <- sample(1:185) #Output of mutCountMatrix earlier puts rows/cols in a certain order - we want this 'random' for better data visualisation
mtx.b <- mtx.b[rowshuffle,]
labels <- amlclust_avg$labels_avg[rowshuffle]

currentAnnotationRow <- data.frame(
  Cluster = factor(labels)
)
rownames(currentAnnotationRow) <- rownames(mtx.b) 

aml_genes <- colnames(mtx.b)[amlclust_avg$varsel_avg == 1]

#Heatmap with selected variables
pheatmap(mtx.b[order(labels),aml_genes], cluster_rows = F,
         color = colorRampPalette(colors = c("white", "black"))(2),
         annotation_row = currentAnnotationRow, cluster_cols = F, show_rownames = F, show_colnames = T,
         legend_breaks = c(0.25, 0.75), legend_labels = c(0, 1))
#Heatmap with all variables
pheatmap(mtx.b[order(labels), sample(1:151)], cluster_rows = F,
         color = colorRampPalette(colors = c("white", "black"))(2),
         annotation_row = currentAnnotationRow, cluster_cols = F, show_rownames = F, show_colnames = F,
         legend_breaks = c(0.25, 0.75), legend_labels = c(0, 1))

```

## ORA analysis
Run ORA analysis and create corresponding plots.
```{r ora}

organism = "org.Hs.eg.db"
BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
library(enrichplot)
library(DOSE)

aml_entrez <- select(org.Hs.eg.db, 
                     keys = aml_genes,
                     columns = c("ENTREZID", "SYMBOL"),
                     keytype = "SYMBOL")

all_aml_entrez <- select(org.Hs.eg.db, 
                         keys = colnames(mtx.b),
                         columns = c("ENTREZID", "SYMBOL"),
                         keytype = "SYMBOL")

```

```{r ora viz}
library(cowplot) #Stitch two plots together
library(ggplot2)

AML_DO <- enrichDO(aml_entrez$ENTREZID, universe = all_aml_entrez$ENTREZID)
dotplot(AML_DO, title = "Disease Ontology ORA", font.size = 10) + xlim(0.65, 1.02) + scale_size_continuous(range  = c(3, 7), 
                                                                                                           limits = c(4, 6), 
                                                                                                           breaks = c(4, 5, 6))
AML_NCG <- enrichNCG(aml_entrez$ENTREZID, universe = all_aml_entrez$ENTREZID)
ncg <- dotplot(AML_NCG, title = "Network of Cancer Genes", font.size = 10)
AML_DGN <- enrichDGN(aml_entrez$ENTREZID, universe = all_aml_entrez$ENTREZID)
dgn <- dotplot(AML_DGN, title = "DisGeNET", font.size = 10) + xlim(0.81, 1.02) + 
  scale_size_continuous(range  = c(3, 7), 
                        limits = c(4, 6), 
                        breaks = c(4, 5, 6))
cowplot::plot_grid(ncg, dgn, labels=LETTERS[1:2], rel_widths=c(.8, .8))


```
